\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

This dissertation describes the design, implementation, and evaluation of an
\textit{Investor Sentiment Dashboard}: an end-to-end analytics system that applies
Natural Language Processing (NLP) and Explainable Artificial Intelligence (XAI)
to the multi-source analysis of public investor sentiment and its relationship to
financial asset price movements.

Three novel contributions are presented, each supported by empirical or
demonstrable results.
First, a \textit{finance-specific emotion taxonomy} layers six domain emotions
(fear, optimism, uncertainty, confidence, scepticism, and mixed) over FinBERT's
softmax probabilities. It combines normalised Shannon entropy, domain lexicon
matching, and financial aspect priors.
This specific combination of signals within a single, architecture-agnostic
scoring function does not appear in the reviewed literature. The module is
deployed as part of the delivered system and annotates the full
150{,}000-record corpus.
Second, a lightweight \textit{Aspect-Based Sentiment Analysis (ABSA)} module
extracts and tags evidence sentences across six financial aspect categories without
requiring fine-tuned sequence models, providing traceable justification for each
headline sentiment label.
Third, \textit{end-to-end XAI integration}: LIME \citep{ribeiro2016lime} is
wired through the full application stack as a REST endpoint and rendered as a
token-level heatmap within the React SPA. Surveys of XAI in finance report
that such explanations typically remain confined to offline notebooks
\citep{yeo2023xai, biswas2024xai}; embedding them directly in a deployed user
interface addresses this deployment gap.
The LIME integration also yielded a concrete diagnostic finding: contextually
neutral risk vocabulary (e.g., ``cautious'', ``inflation'') is systematically
pushed toward the negative class. This identifies the mechanism behind
FinBERT's neutral recall deficit and points to a targeted fine-tuning strategy.

To underpin these contributions, \textbf{FinBERT} \citep{araci2019finbert} is
deployed as the core classifier on text aggregated from three distinct channels:
Reddit (via PRAW), X/Twitter (via the official X API v2 with a historical CSV
fallback for reproducibility), and financial news (via NewsAPI).
On a purpose-built, 200-sample evaluation benchmark reflecting the real
distribution of multi-source financial text, FinBERT achieves \textbf{82.0\%}
accuracy and \textbf{80.3\%} macro F1-score, \textbf{outperforming a
keyword-based baseline by 13.5 and 12.4 percentage points respectively}, a
difference that is statistically significant under a paired McNemar test
($\chi^2_{cc} = 9.52$, $p = 0.002$).
The most striking individual gain is on the negative class, where FinBERT
recall improves from 47.1\% to 94.3\%. This 47.2 percentage-point improvement
illustrates the inherent limitation of lexicon-based methods on implicit
financial phrasing.

In addition to the three primary contributions, a \textit{sentiment-price
correlation engine} provides Pearson and Spearman correlation, configurable lag
analysis, Granger causality testing \citep{granger1969}, rolling correlation, and
market-adjusted metrics within a single interactive interface. A demonstrative
run against the live corpus across a nine-ticker panel is reported in the
evaluation chapter: same-day Pearson coefficients span $r \in [-0.10, +0.24]$,
with NVDA, AAPL, and AMZN individually significant at $\alpha = 0.05$
(uncorrected) and SPY producing the expected near-zero null for a diversified
index tracker. No Granger test reaches significance at $\alpha = 0.05$ on the
one- or two-day lags examined, which is reported transparently rather than
selectively.

Overall, the project demonstrates that multi-source, domain-adaptive sentiment
analysis can be deployed transparently and at scale within a single academic
system. The finance-emotion taxonomy, ABSA enrichment layer, and end-to-end
XAI integration together map onto a coherent set of research contributions at
the intersection of NLP, behavioural finance, and explainable AI. Limitations,
including the modest size of the Twitter subcorpus and the absence of a
full multi-annotator human panel for benchmark labelling (partially addressed
by LLM cross-annotation across all 200 samples and a supplementary
single-volunteer human annotator check on a 100-sample subset), are recorded
transparently in the evaluation and discussion chapters.
